---
title: 'Data Types'
author: "Naomi Tague"
date: "January, 2023"
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)

library(tidyverse)
library(purrr)

```

# <span style="color:blue"> Data types 

All programming languages use data-types, or structures to hold information

* integer
* floating point/ real / numeric
* character 
* string

Often data types are multi-dimensional 
Some useful ones in R

* vector
* matrix
* data frame
* tibble
* factors
* lists

Programming often involves selecting and building data structures. Like the **res** matrix we built last class to hold the results from our **for** loop

Good data structures are

* as simple as possible
* easy to understand (readable names)
* easy to manipulate 
* easy to visualize

# <span style="color:blue"> Factors \span

something that has different **classes** or **groups**
useful for doing calculations with categories

Here's an example:

First lets look at a standard numeric vector

```{r} 
a = c(1.3, 1, 4, 1.3, 22)
# compute the mean
mean(a)
```

What if **a** is a factor

What do commands like **mean** do
```{r} 
a = as.factor(a)
# compute the mean
mean(a)

#why? lets look
a




```

We can use **summary** with factors to get frequencies in each category (or “level” )



```{r fishes}

# create vector of possible fish 
possible.fish = c("salmon","steelhead","shark","tuna","cod")

# we can use sample to simulate a random recording of catch by fisherman, lets say we pick 20 fish from the net

catch1 = base::sample(possible.fish, size=20, replace=T)
# because possible.fish was a factor catch1 will be a factor
catch1

summary(catch1)
# if we want summary to be more useful - make this a factor
catch1 = as.factor(catch1)


# to quickly get frequencies of different fish and to plot 
summary(catch1)
plot(catch1, col="blue")


# we can also use summary to explore and return information about the distribution
# mean frequency of a particular type of fish
mean(summary(catch1))

# maximum frequency
max(summary(catch1))

# which fish was most frequently caught
which.max(summary(catch1))

#to get just the name 
names(which.max(summary(catch1)))

# use results for creating text
# sprintf creates a string %s mean use what ever is after the , to fill in a string
plottitle=sprintf("We like %s", names(which.max(summary(catch1))))

ggplot(data.frame(catch1=catch1), aes(catch1, fill=catch1))+geom_histogram(stat="count")+labs(title=plottitle)

# you can also add numbers to the string
plottitle=sprintf("We mostly caught %s \n max catch(%d)", names(which.max(summary(catch1))), max(summary(catch1)))
ggplot(data.frame(catch1=catch1), aes(catch1, fill=catch1))+geom_histogram(stat="count")+labs(title=plottitle)

#How do you figure out the rarest fish in our simulated ocean


```

# Aside **sprintf**

some useful syntax if you want to generate strings

* **%s** replace with a string
* **%d** replace with an integer value
* **%f** replace with a real value
* **%4.1f** replace with a real value with 4 digist, two after decimal
* **\\n** add a line return

Try it

make a string with **sprintf** and add to plot title


# <span style="color:blue"> Functions with factors 

Lets generate a function that makes use of categorical data
species diversity is a good example

"Simpson's Index (D) measures the probability that two individuals randomly selected from a sample will belong to the same species 

Value is between 0 and 1, with lower values associated with *lower* diversity

See 
[Simpson Biodiversity](http://www.countrysideinfo.co.uk/simpsons.htm)


```{r diversity, echo=TRUE}

source("R/compute_simpson_index.R")
compute_simpson_index


possible.fish = as.factor(c("salmon","steelhead","shark","tuna","cod"))
# simulate a random recording of catch by fisherman


# note here is answer to above challenge
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)
# lets create a test case that should have low diversity, by repeating the same thing
catch2 = c(rep("salmon", times=10), rep("cod", times=10))

compute_simpson_index(catch1)
compute_simpson_index(catch2)

```


Sometimes you don't want factors and R thinks something should be
How to change back? **as.numeric** makes sense ...but



```{r, echo=TRUE}

a = as.factor(c(1.3, 1, 4, 1.3, 22))
#sum(a)

# try to make a numeric version from the factor
b = as.numeric(a)
sum(b)
b

# better
b = as.character(a)
b = as.numeric(b)
b
sum(b)
```

#  <span style="color:blue"> Returning multiple things from a function \span

In R, to do this we use LISTS

* Lists are the most “informal” data structures in R
* List are really useful for keeping track of and organizing groups of things that are not all the same
* A list could be a table where number of rows is different for each column
* A list can have numeric, character, factors all mixed together
* List are often used for returning more complex information from function (e.g. lm)

```{r introlist, echo=TRUE}

# make a list
sale = list(id=2, quality="high", contents=c("apple","cherry"), cost=c(4,5))
sale

#ways to access elements
sale$id
sale$what

# you can also access by list item number
# the [x] denotes the xth item in the list
sale[[3]]
sale[["contents"]]


# how do you get the second element in the vector that has the contents
# there are two ways


# add to a list
sale$location = "Farmers market"
sale
# or remove
sale$location = NULL
sale

# some tricky things
# correct accessing items in list
sale$cost
sale[[4]]

# works but
#sale[4]


sum(sale$cost)
sum(sale[[4]])

```


# Using lists to return multiple items from a function

We can use *lists* to return multiple,diverse pieces of information from our functions
Lets start with diversity - many be want to know a bit more about the dataset

* Simpson diversity
* most frequent species
* number of distinct species



```{r diversitylist, echo=TRUE}

# repeat with a list
source("R/computediversity.R")

computediversity

computediversity(catch1)
computediversity(catch2)
```


