---
title: 'Error Checking '
author: "Naomi Tague"
date: "January, 2021"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
library(tidyverse)
```



# Review

* We are using functions as a way to design modular programs (that might involve many functions that are linked together)

* We always start with clearly defininng our goal for the whole program and each piece (and diagramming this if needed for programs made of multiple functions)

* Always write your function in a text editor and then copy into R

* By convention we name files with functions in them by the name of the function.R
e.g. **power_gen_orig.R**

* you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

* keep organized by keeping all functions in a subdirectory called *R*

*  Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”




```{r}

source("R/power_gen_orig.R")
```


# Multiple inputs 

For a function with more than one input  R can handle multiple values


**IF** 

* all (multiple) inputs have the same number of values 
* all other inputs or parameters have a single value


For other cases, additional instructions would need to be included in the function to help R understand what to do


Consider our reservoir function  ( **R/power_gen_orig.R**)

```{r sampling2}

#source("R/power_gen_orig.R")

power_gen_orig= function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  
  # calculate power
  result = rho * height * flow * g * Keff
  return(result)
}

# single
power_gen_orig(height=2, flow=2)

# multiple heights
# generate a structure to store results
# consider all heights from 1 to 100

resv_results = data.frame(height=seq(from=1, to=100))

# compute power generated for EACH height
resv_results$power = power_gen_orig(resv_results$height, flow=2)

# plot
ggplot(resv_results, aes(height, power))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# random combinations of height and flow
# use rnorm to sample from a normal distribution
resv_results$height = rnorm(mean=10, sd=2, n=100)
resv_results$flow = rnorm(mean=20, sd=1, n=100)
# compute power generated for EACH height
resv_results$power = power_gen_orig(resv_results$height, flow=resv_results$flow)

# plot
ggplot(resv_results, aes(height, power, col=flow))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# Unequal lengths - this will "work" but inputs get recycled unclear what is happening
# NOT recommended

eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=100)
resv_result = power_gen_orig(eheight, eflow)

# better
eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=50)
resv_result = power_gen_orig(eheight, eflow)

# another example of the problem of unequal length inputs
power_gen_orig(eheight, eflow, g=c(9,10,11))

#better
g = runif(min=8, max=10, n=50)
resv_result = power_gen_orig(eheight, eflow, g)
ggplot(resv_results, aes(height, power, col=flow))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")


```

# Error Checking

What will your function do if user gives you garbage data, imagine that you have a function that computes microbial growth rates with temperature

* Two options

<span style="color:skyblue">
1. error-checking (nice)
</span>

 if temperature < -100 or > 100
 
*  set results to NA
*  or exit function with a warming
 
<span style="color:skyblue"> 
2. assume user reads the contract :) return unrealistic values
</span>

if input -999.0, will still try to output growth rate
 
 
 Error-checking is helpful if you are going to build a model made up of many functions- why?
 

# How to add error checking

Consider this example (I'm going to put the function definition inline for now so you can easily see it - later we will save to our R subdirectory for reuse)

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
 # an alterative is to exit the function prematurely 
  if (flow < 0) return("Flow cannot be less than 0")
  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)
```


While this works for a case with one height and one flow
It will have problems if we use multiple heights - (a later topic)

A better implementation if you think the function might be run for multiple heights - YES you can do that

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   height = ifelse( (height < 0), NA, height)
  
 # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)

  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)


# note that this works (calculates power for each height)
power_gen(height=c(1,4,5), flow=2)

# and this works (calculates power for each flow rate)
power_gen(height=4, flow=c(2,4,5))

# and this calculates power for each combination of flow and height
power_gen(height=c(4,5,6), flow=c(2,4,6))

# but this doesn't work - why?
power_gen(height=c(4,5,6, 8, 9, 20), flow=c(2,4,6))

# we could add error checking to test for this

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  
  # make sure we have same number of flow and height's so we can match them
  if (length(height) != length(flow)) return(" number of heights is different from number of flows")
  # make sure inputs are positive and if not set as NA so result will be NA
   height = ifelse( (height < 0), NA, height)
  
 # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)

  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(height=c(4,5,6, 8, 9, 20), flow=c(2,4,6))
power_gen(height=c(4,5,6), flow=c(2,4,6))
```

See **power_gen.R** for file with error checking

Error checking is good practice - and combinded with good documentation - makes programming much much easier!




