---
title: "Sampling_example.Rmd"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
library(tidyverse)

```
Lets imagine you've called your auto power function "autopower" and saved
it in the R directory of your workspace as 
**autopower.R**

You could use any name you want - you will just need to change the name in the example code below

# Generating data for the function and iterating over that data



Example use: Imagine we want to see how much power is generated given scenario where we know the mean and standard deviation of vehicles speeds

Here's how we would do this

Note documentation in the R markdown - and use of variables names 

```{r sampling}

source("R/autopower.R")

# generate sample speeds from a distribution
nsample = 100
speeds = rnorm(mean=25, sd=4, n=nsample)

# Step 1  create data frame to store results 
# how many simulations, what do you want to keep

#create a dataframe that has rows for each model run
# columns for height, flowrate and power estimate
results = data.frame(speed=speeds, power=NA)

head(results)

# if you only have one input and everything else is the same, R is smart enough
# to create a set of outputs
results$power = autopower(V=speeds, A=25, m=20000)

# ALWAYS look at your results to make sure it make sense
ggplot(results, aes(speed, power/1000))+geom_point()+labs(x="speed in m/s", y="power in kW")
ggplot(results, aes(x="Across Speed",y=power/1000))+geom_boxplot()+labs(y="power in kW", x="")

```
