---
title: 'Functions'
author: "Naomi Tague"
date: "January, 2021"
output:
  slidy_presentation: default
  ioslides_presentation: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
library(tidyverse)
```


## STEPS IN Program Design

1. Clearly define your goal in words/figures as precisely as possible what do you want your program to do
      + decide on modules
      + for each module, write a contract - inputs, outputs and what it will do
    

2. Implement AND document as you go
3. Test
4. Refine
5. Test again
6. Create User Documentation and Distribute


---

## A simple function example

Contract

*Input*: Reservoir height (m) and flow rate (m3/sec) (as a single value in time or a time series)

*Output*: Instantaneous power generation (W/s)  (as a single value or value for each reservoir height and flow rate combination)

*Parameters*: K_{Eff} reservoir efficiency, ρ (density of water), g (acceleration due to gravity)

What the function will do - compute power using the following equation - *body* of the function 

$P = ρ * h * r * g * K_{Eff}$

P is Power in watts, ρ is the density of water (~1000 $kg/m^3$), h is height in meters, r is flow rate in cubic meters per second, g is acceleration due to gravity of 9.8 $m/s^2$, $K_{Eff}$ is a coefficient of efficiency ranging from 0 to 1.

---

# Implementing Functions in R

Format for a basic function in R


\#' documentation that describes inputs, outputs and what the function does

FUNCTION NAME = function(inputs, parameters) {

body of the function (manipulation of inputs)

return(values to return)

}

# Implementing Functions in R

In R, inputs and parameters are treated the same; but it is useful to think about them separately in designing the model - collectively they are sometimes referred to as arguments

ALWAYS USE Meaningful names for your function, its parameters and variables calculated within the function


* In R we combine inputs/parameters in the first line of the function definition

* we can provide "default" values that can be overwritten by the function user

* Body includes all calculations between { and }
  + anything inside the body is not "seen" by the workspace so you can re-use variables
  
* *return* tells R what the output is

# Here's our example

```{r eval=TRUE, echo=TRUE  }

# function definition
power_gen_orig = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
result = rho * height * flow * g * Keff
return(result)
}
```

---

# Using the function and scoping

To use the function we substitute "actual" values for each argument (parameter and inputs)
(unless there are defaults)

R assumes that arguments are in the order that they were specified in the function definition - unless
you refer to them by name

```{r eval=TRUE, echo=TRUE  }

power_gen_orig(20,1)

power_gen_orig(height=20, flow=1)

power_gen_orig(flow=1, height=20)

# change the defaults
# what if efficiencies is less

power_gen_orig(20,1, Keff=0.2)

```

---

# Scoping


The scope of a variable in a program defines where it can be “seen” Variables defined inside a function cannot be “seen” outside of that function

There are advantages to this - the interior of the building block does not ‘interfere’ with other parts of the program

See below

```{r eval=TRUE, echo=TRUE  }

# specifying function name without () shows code
# try it for internal R functions
power_gen_orig

# use the function and save the results
reservoir1 = power_gen_orig(height=4, flow=7)

reservoir2 = power_gen_orig(height=4, flow=7, Keff=0.2)

# note that internal variables are not defined so these R statements will FAIL
height
Keff
results

# only the outputs are in your workspace
reservoir1
reservoir2


```

---

# Some conventions (helpful later in the course)

* Always write your function in a text editor and then copy into R

* By convention we name files with functions in them by the name of the function.R
e.g. **power_gen_orig.R**

* you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

* keep organized by keeping all functions in a subdirectory called *R*

* Include comments both in the body of the function and at the top  - there is a particular style of adding comments to the top of the function that we will use later to generate automatic documentation - so useful to start to follow it, more or less, now - see **power_gen.R** for the style



*  Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”


