---
title: 'Testing'
author: "Naomi Tague"
date: "January, 2020"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
#knitr::opts_knit$set(root.dir = '/Users/christina/Desktop/Courses.local/ESM262n/docs/packages')
library(tidyverse)
library(devtools)
library(testthat)

```



# <span style="color:orange">Testing in R</span>

In R, *testthat* library 
install and add this library

If you are in the working directory where you’ve stored the files for your project you can use *testthat* to automate testing of your functions

* need devtools and “testthat” libraries

A few R commands that are useful

* test_dir(“name”) :runs all tests in the “name” subdirectory  (all files beginning with the word “test”
* test_file(“name”): runs all  the tests in a file called “name”

Testing in R uses "expectation"

# <span style="color:orange">Expectation</span>


tests you can use to make sure your code is working the way you think it should be working

* expect_true(condition)
* expect_equal(value, value)
* expect_match(string1, string2)
* expect_null(value)
* expect_length(object, n)



```{r testingsimple, eval=TRUE, error=TRUE}


# runs a test on a statement and checks to see if it is correct
# this one should work
expect_equal(2+2, 4)
# this one should 'throw and error'
expect_equal(2+5, 4)

# we can also use expect_true
# works with variables two
a = 200
expect_true(a > 2 )
a = 0
expect_true(a > 2 )

# matches
fish1 = "salmon"
fish2 = "trout"
expect_match(fish1, fish2)

# length
a = seq(from=1, to=10)
expect_length(a, 10)
expect_length(a, 11)

# We can use expectation with functions 
# and test just to make sure they are working as expected
source("R/compute_NPV.R")
futurevalue = 100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)
futurevalue = -100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)

# what if function returns a list?
# we can use the buit in summary function to see how this works
testdata = rnorm(mean=25, sd=3, n=100)
summary(testdata)
# it is actually returning a list, but what are the names?
names(summary(testdata))
# now we can specify them
summary(testdata)["Max."]
# and use in a test to make sure this function works -...hmm
# well the max should be greater than the min right?

expect_true((summary(testdata)["Max."]-summary(testdata)["Min."]) > 0)

# lets create a function with an error 
# imagine total recharge of water to soil after rain (in mm) is daily rainfall minus interception (in mm with default of 1mm) lost on any given day
# lets define a function that returns total rainfall and total recharge given an input of
# rainfall and then a value for interception

effective_rain=function(rain, interception_loss=1) {
  
  # easy part - get total rain
  totalrain = sum(rain)
  
  # interception happens everyday so we need a new vector
  recharge = rain-interception_loss
  totalrecharge = sum(recharge)
  
  return(list(totalrain=totalrain, totalrecharge=totalrecharge))}


# lets do some simple test
# if it never rains for 730 days then both should be zero
raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrain, equals(0))
# so far so good, now for recharge

raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrecharge, equals(0))

# what went wrong???
```
